---
title: "Life Contingencies with tidyactuarial"
author: "Julian Fajardo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Life Contingencies with tidyactuarial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(tidyactuarial)
library(dplyr)
library(tibble)
```

## From financial value to actuarial value

Life-contingency mathematics adds one element to the financial framework: a payment may depend on survival or death. The financial discount factor still determines how money is moved through time, but it must now be combined with a probability describing whether the payment occurs.

For a life aged $x$, let ${}_tp_x$ denote the probability of surviving $t$ years. If

$$
v = \frac{1}{1+i},
$$

then a unit payment at time $t$, made only if the life survives, has actuarial present value

$$
{}_tE_x
=
v^t\,{}_tp_x.
$$

This quantity is the **pure endowment factor**. It is the simplest bridge between financial mathematics and life contingencies.

The same idea extends to annuities, insurance benefits, premiums, and reserves. The objective of this vignette is to keep that progression visible:

$$
\text{life table}
\longrightarrow
\text{survival and death}
\longrightarrow
\text{actuarial present values}
\longrightarrow
\text{premium}
\longrightarrow
\text{reserve}.
$$

The mathematics is written first. `tidyactuarial` is then used to reproduce and audit the calculation.

## A small life table

For integer ages, a life table records the number $l_x$ expected to survive to exact age $x$. The basic one-year quantities are

$$
p_x
=
\frac{l_{x+1}}{l_x},
\qquad
q_x
=
1-p_x
=
\frac{l_x-l_{x+1}}{l_x}.
$$

More generally, for an integer duration $t$,

$$
{}_tp_x
=
\frac{l_{x+t}}{l_x},
\qquad
{}_tq_x
=
1-{}_tp_x.
$$

We begin with a deliberately small table so every result can still be checked by hand.

```{r life-table}
lt <- lifetable(
  x = 60:66,
  lx = c(
    100000,
    99000,
    97500,
    95500,
    93000,
    90000,
    86000
  ),
  close = FALSE,
  frac = "UDD"
)

lt
```

For example,

$$
{}_3p_{60}
=
\frac{l_{63}}{l_{60}}
=
\frac{95500}{100000}
=
0.9550,
$$

and therefore

$$
{}_3q_{60}
=
1-0.9550
=
0.0450.
$$

The same quantities can be calculated directly with `t_px()` and `t_qx()`.

```{r survival-probabilities}
survival_audit <- tibble(
  duration = 0:5
) |>
  mutate(
    survival = t_px(
      lt = lt,
      x = 60,
      t = duration,
      frac = "UDD"
    ),
    death = t_qx(
      lt = lt,
      x = 60,
      t = duration,
      frac = "UDD"
    )
  )

survival_audit
```

The pipe does not replace the actuarial formulas. It simply lets several durations be evaluated in one transparent workflow.

## Actuarial discounting: the pure endowment

Suppose the annual effective interest rate is $i=5\%$. A payment of 1 at time 5, conditional on survival of $(60)$ to age 65, has value

$$
{}_5E_{60}
=
v^5\,{}_5p_{60}.
$$

Since

$$
{}_5p_{60}
=
\frac{90000}{100000}
=
0.9000,
$$

we obtain

$$
{}_5E_{60}
=
(1.05)^{-5}(0.9000)
\approx
0.7052.
$$

The manual calculation and the package calculation should agree.

```{r pure-endowment}
i <- 0.05
v <- 1 / (1 + i)

pure_endowment_manual <- v^5 * 90000 / 100000

pure_endowment_package <- t_Ex(
  lt = lt,
  x = 60,
  t = 5,
  i = i,
  frac = "UDD"
)

tibble(
  method = c(
    "Manual",
    "tidyactuarial"
  ),
  value = c(
    pure_endowment_manual,
    pure_endowment_package
  )
)
```

Financial discounting and survival are not competing ideas. They are the two components of the same actuarial present value.

## A temporary life annuity-due

Consider a 5-year temporary life annuity-due on $(60)$ with unit annual payments. Payments occur at times

$$
0,1,2,3,4,
$$

provided the life is alive at each payment date. Its actuarial present value is

$$
\ddot a_{60:\overline{5}|}
=
\sum_{k=0}^{4}
v^k\,{}_kp_{60}.
$$

Using the life table directly,

$$
\ddot a_{60:\overline{5}|}
=
1
+
v\frac{l_{61}}{l_{60}}
+
v^2\frac{l_{62}}{l_{60}}
+
v^3\frac{l_{63}}{l_{60}}
+
v^4\frac{l_{64}}{l_{60}}.
$$

```{r life-annuity}
survival_0_4 <- lt$lx[1:5] / lt$lx[1]

annuity_manual <- sum(
  v^(0:4) * survival_0_4
)

annuity_package <- annuity_x(
  lt = lt,
  x = 60,
  i = i,
  n = 5,
  timing = "due",
  frac = "UDD"
)

tibble(
  method = c(
    "Manual",
    "tidyactuarial"
  ),
  apv = c(
    annuity_manual,
    annuity_package
  )
)
```

The annuity value is approximately $4.4173$. This is not the value of five certain payments. Each future payment is discounted financially and weighted by the probability that the life remains alive.

## A temporary life insurance

Now consider a 5-year term insurance on $(60)$ with benefit 1 payable at the end of the year of death.

The probability that death occurs in policy year $k+1$ is

$$
{}_kp_{60}q_{60+k}
=
\frac{l_{60+k}-l_{61+k}}{l_{60}},
$$

for $k=0,\ldots,4$. Therefore,

$$
A^{1}_{60:\overline{5}|}
=
\sum_{k=0}^{4}
v^{k+1}
{}_kp_{60}q_{60+k}.
$$

```{r term-insurance}
death_0_4 <- (
  lt$lx[1:5] - lt$lx[2:6]
) / lt$lx[1]

insurance_manual <- sum(
  v^(1:5) * death_0_4
)

insurance_package <- insurance_x(
  lt = lt,
  x = 60,
  i = i,
  n = 5,
  type = "term",
  benefit = 1
)

tibble(
  method = c(
    "Manual",
    "tidyactuarial"
  ),
  apv = c(
    insurance_manual,
    insurance_package
  )
)
```

For a benefit of 100,000, the actuarial present value of benefits is simply

$$
100000
A^{1}_{60:\overline{5}|}.
$$

```{r insurance-benefit}
benefit <- 100000

insurance_x(
  lt = lt,
  x = 60,
  i = i,
  n = 5,
  type = "term",
  benefit = benefit
)
```

## An actuarial identity as an audit

The temporary insurance, temporary annuity-due, and pure endowment satisfy

$$
A^{1}_{x:\overline{n}|}
=
1
-
d\,\ddot a_{x:\overline{n}|}
-
{}_nE_x,
$$

where

$$
d
=
\frac{i}{1+i}.
$$

This identity is useful because the three quantities were obtained through different calculations.

```{r identity-audit}
d <- i / (1 + i)

identity_value <- 1 -
  d * annuity_package -
  pure_endowment_package

tibble(
  direct_insurance = insurance_package,
  identity_value = identity_value,
  difference = insurance_package - identity_value
)
```

The difference should be zero up to numerical precision. The point of the audit is not to replace the theory with a test; it is to verify that the separate implementations preserve the same actuarial identity.

## Net premium by the equivalence principle

Suppose the 5-year term insurance has benefit 100,000 and level annual premiums payable in advance while the insured is alive, for at most five years.

At issue, the equivalence principle requires

$$
\operatorname{APV}(\text{premiums})
=
\operatorname{APV}(\text{benefits}).
$$

If $P$ is the annual net premium,

$$
P\,
\ddot a_{60:\overline{5}|}
=
100000
A^{1}_{60:\overline{5}|},
$$

so

$$
P
=
\frac{
100000A^{1}_{60:\overline{5}|}
}{
\ddot a_{60:\overline{5}|}
}.
$$

The manual premium is therefore approximately 1,912.47.

```{r net-premium}
premium_manual <- benefit *
  insurance_manual /
  annuity_manual

premium_package <- premium_x(
  lt = lt,
  x = 60,
  i = i,
  type = "term",
  benefit = benefit,
  n = 5,
  k = 1,
  n_prem = 5,
  timing = "due",
  output = "value"
)

tibble(
  method = c(
    "Manual",
    "tidyactuarial"
  ),
  premium = c(
    premium_manual,
    premium_package
  )
)
```

For an audit-oriented result, `premium_x()` can also return the components of the equation of equivalence.

```{r premium-summary}
premium_x(
  lt = lt,
  x = 60,
  i = i,
  type = "term",
  benefit = benefit,
  n = 5,
  k = 1,
  n_prem = 5,
  timing = "due",
  output = "summary"
)
```

This distinction is useful in actuarial work: a single value answers the pricing question, while the summary exposes the quantities needed to audit the result.

## Reserves: valuing the remaining contract

A premium determined at issue does not remain the appropriate measure of the insurer's obligation at later policy durations. Conditional on the policy still being in force at duration $t$, the prospective reserve is

$$
{}_tV
=
\operatorname{APV}_t
(\text{future benefits})
-
P\,
\operatorname{APV}_t
(\text{future premiums}).
$$

Consider duration $t=2$, immediately before the premium due at that duration. The insured is then age 62 and the contract has three years remaining.

The remaining premium annuity is

$$
\ddot a_{62:\overline{3}|},
$$

and the remaining term-insurance value is

$$
A^{1}_{62:\overline{3}|}.
$$

Thus,

$$
{}_2V
=
100000
A^{1}_{62:\overline{3}|}
-
P
\ddot a_{62:\overline{3}|}.
$$

```{r reserve-manual}
survival_62 <- lt$lx[3:5] / lt$lx[3]

annuity_62 <- sum(
  v^(0:2) * survival_62
)

death_62 <- (
  lt$lx[3:5] - lt$lx[4:6]
) / lt$lx[3]

insurance_62 <- sum(
  v^(1:3) * death_62
)

reserve_2_manual <- benefit *
  insurance_62 -
  premium_package *
  annuity_62

reserve_2_manual
```

The same contract can be valued at several policy durations with `reserve_x()`.

```{r reserve-schedule}
reserve_schedule <- reserve_x(
  lt = lt,
  x = 60,
  i = i,
  type = "term",
  n = 5,
  benefit = benefit,
  P = premium_package,
  k = 1,
  n_prem = 5,
  timing = "due",
  t = 0:5,
  method = "prospective",
  output = "summary"
)

reserve_schedule
```

For this annual fully discrete contract, the recursive method provides a second computational route.

```{r reserve-recursive}
reserve_x(
  lt = lt,
  x = 60,
  i = i,
  type = "term",
  n = 5,
  benefit = benefit,
  P = premium_package,
  k = 1,
  n_prem = 5,
  timing = "due",
  t = 0:5,
  method = "recursive",
  output = "summary"
)
```

The prospective and recursive approaches answer the same actuarial question from different directions. Agreement between them is therefore a substantive audit, not merely a software check.

## The same contract as a pipe workflow

The direct calculations above make each actuarial quantity explicit. Once the contract is understood, the same specification can be represented as a short workflow.

```{r contract-pipeline}
contract <- life_contract(
  lt = lt,
  lives = "single",
  x = 60,
  i = i
) |>
  add_insurance(
    type = "term",
    benefit = benefit,
    n = 5
  ) |>
  add_premium_schedule(
    k = 1,
    n_prem = 5,
    timing = "due"
  )

contract |>
  premium_x(
    output = "summary"
  )
```

This is where a pipe is useful: it follows the actuarial construction of the contract itself.

1. Define the life and financial basis.
2. Add the insurance benefit.
3. Add the premium-payment schedule.
4. Apply the equivalence principle.

The pipe is not used to hide the mathematics. It records the contract in the same order in which the actuary specifies it.

## Moving to a benchmark life table

The small table above was useful because every quantity could be reproduced manually. For realistic calculations, the same functions work with larger tables.

`tidyactuarial` includes `soa08lt`, a tidy version of the SOA Illustrative Life Table intended for reproducible actuarial examples and benchmark calculations.

```{r soa-benchmark}
data("soa08lt")

soa08lt |>
  filter(
    x >= 60,
    x <= 65
  ) |>
  select(
    x,
    lx,
    dx,
    qx,
    px
  )
```

For example, a whole-life annuity-due and a whole-life insurance at age 65 can be valued without changing the mathematical language of the previous sections.

```{r soa-values}
tibble(
  quantity = c(
    "Whole-life annuity-due",
    "Whole-life insurance, benefit 100000"
  ),
  value = c(
    annuity_x(
      lt = soa08lt,
      x = 65,
      i = 0.05,
      timing = "due"
    ),
    insurance_x(
      lt = soa08lt,
      x = 65,
      i = 0.05,
      type = "whole",
      benefit = 100000
    )
  )
)
```

The difference is scale, not principle.

## Final actuarial reading

The complete workflow can now be summarized without introducing new formulas.

A life table provides survival and death probabilities. Financial discounting moves each possible payment to the valuation date. Their combination produces actuarial present values. The equivalence principle converts those values into a premium, and the same valuation principle applied at a later duration produces a reserve.

In symbols,

$$
v^t
\quad+\quad
{}_tp_x
\quad\longrightarrow\quad
{}_tE_x,
$$

and, more generally,

$$
\text{discounting}
+
\text{mortality}
\longrightarrow
\text{APV}
\longrightarrow
\text{premium}
\longrightarrow
\text{reserve}.
$$

`tidyactuarial` keeps this progression reproducible while preserving the actuarial structure of the calculation. The package is most useful after the contract has been stated correctly: it does not replace the equation of value, the mortality model, or the timing convention; it makes them easier to calculate, compare, and audit.
