---
title: "Bonds, Duration, Convexity, and Immunization"
author: "Julian Mauricio Fajardo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bonds, Duration, Convexity, and Immunization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

library(tidyactuarial)
```

## Introduction

A bond converts a contractual sequence of future payments into a value today. For a conventional fixed-coupon bond, the main inputs are the coupon payments, the redemption value, the payment dates, and the yield used to discount the cash flows.

The coupon rate and the yield play different roles. The coupon rate is contractual and determines the periodic payment. The yield is the rate used to value those payments. A change in market yield does not modify the contractual coupon; it modifies the present value of the remaining cash flows and therefore the bond price.

This vignette starts from the usual actuarial equations and then uses `tidyactuarial` to reproduce and audit the calculations. The mathematical structure is established first; the software is used afterward as a computational tool.

## Bond cash flows and the value equation

Consider a conventional bond with face value $F$, coupon rate $r$ per coupon period, redemption value $C$, effective yield $i$ per coupon period, and $n$ remaining coupon periods. The periodic coupon is $Fr$.

Taking the valuation date as time $0$, the future cash flows are

$$
Fr,\; Fr,\; \ldots,\; Fr+C.
$$

Therefore,

$$
P
=
Fr(v+v^2+\cdots+v^n)+Cv^n,
$$

where $v=(1+i)^{-1}$. Since the coupon payments form an annuity-immediate,

$$
P
=
Fr\,a_{\overline{n}|i}
+
Cv^n.
$$

Equivalently,

$$
P
=
Fr\frac{1-(1+i)^{-n}}{i}
+
C(1+i)^{-n}.
$$

This is the ordinary equation of value applied to a level stream of coupons plus a single redemption payment. The coupon rate, the yield, and the number of periods must refer to the same time unit.

## Valuing a coupon bond

Consider a bond with

$$
F=C=1{,}000{,}000,
$$

a nominal annual coupon rate of $9\%$ payable semiannually, and six years to maturity. Suppose the investor requires a nominal annual yield of $10\%$ convertible semiannually.

The coupon rate per half-year is

$$
r=\frac{0.09}{2}=0.045,
$$

so the semiannual coupon is $45{,}000$. The effective yield per half-year is $i=0.05$, and there are $12$ remaining coupon periods. Hence,

$$
P
=
45{,}000a_{\overline{12}|0.05}
+
1{,}000{,}000(1.05)^{-12}
=
955{,}683.74.
$$

Only after constructing the equation of value do we move to the computational representation.

```{r bond-cash-flows}
face_value <- 1000000
redemption_value <- 1000000
annual_coupon_rate <- 0.09
years_to_maturity <- 6
coupons_per_year <- 2

bond_flows <-
  bond_cash_flows(
    face = face_value,
    c = annual_coupon_rate,
    n = years_to_maturity,
    k = coupons_per_year,
    R = redemption_value
  )

bond_flows
```

`bond_cash_flows()` represents the contractual payments. It does not determine the bond price. The valuation basis is introduced separately.

```{r bond-price}
semiannual_yield <- 0.05

bond_value <-
  bond_price(
    face = face_value,
    c = annual_coupon_rate,
    n = years_to_maturity,
    k = coupons_per_year,
    y_effective_per_period = semiannual_yield,
    R = redemption_value
  )

bond_value
```

The package result reproduces the analytical value. The comparison verifies that the computational implementation follows the actuarial equation established independently.

## Yield to maturity as an inverse problem

Suppose the market price of the same bond is now

$$
P=935{,}000.
$$

The YTM is the rate $i$ satisfying

$$
935{,}000
=
45{,}000\frac{1-(1+i)^{-12}}{i}
+
1{,}000{,}000(1+i)^{-12}.
$$

For this bond,

$$
f(0.05)>0
\qquad\text{and}\qquad
f(0.06)<0,
$$

so the root lies in

$$
5\%<i<6\%.
$$

Since the price of a conventional bond with positive future cash flows is strictly decreasing in the yield, the solution is unique in this interval. Numerically,

$$
i\approx5.2435\%
$$

per half-year.

```{r bond-ytm}
market_price <- 935000

ytm_result <-
  bond_ytm(
    P = market_price,
    face = face_value,
    c = annual_coupon_rate,
    n = years_to_maturity,
    k = coupons_per_year,
    R = redemption_value,
    interval = c(0.05, 0.06)
  )

semiannual_ytm <-
  ytm_result |>
  dplyr::pull(i_period) |>
  dplyr::first()

ytm_result
```

The interval does not come from the software. It is identified from the equation of value before the numerical solver is used.

## Book value

Let $BV_t$ denote the book value at coupon date $t$. Prospectively,

$$
BV_t
=
Fr\,a_{\overline{n-t}|i}
+
Cv^{n-t}.
$$

Retrospectively,

$$
BV_t
=
P(1+i)^t
-
Fr\,s_{\overline{t}|i}.
$$

These are two representations of the same financial equation.

For the bond purchased at $935{,}000$, after three years, or six semiannual coupon periods,

$$
BV_6
\approx
962{,}555.82.
$$

```{r bond-book-value}
valuation_time <- 3

book_value <-
  bond_book_value(
    face = face_value,
    c = annual_coupon_rate,
    n = years_to_maturity,
    t = valuation_time,
    k = coupons_per_year,
    y_effective_per_period = semiannual_ytm,
    R = redemption_value
  )

book_value
```

The recurrence

$$
BV_t=BV_{t-1}(1+i)-Fr
$$

shows how premiums and discounts are recognized through time. Book value should not be confused with market value: the former follows the yield basis adopted for amortization, whereas the latter uses the yield currently required by the market.

## Callable bonds

A callable bond gives the issuer the right to redeem the bond before maturity under specified contractual conditions. The investor does not control the final redemption date.

Suppose scenario $s$ redeems the bond after $n_s$ coupon periods for $C_s$. At target yield $i$,

$$
P_s(i)
=
Fr\,a_{\overline{n_s}|i}
+
C_sv^{n_s}.
$$

To guarantee at least the target yield under every admissible scenario, the purchase price must satisfy $P\le P_s(i)$ for every $s$. Hence,

$$
P_{\max}(i)
=
\min_s P_s(i).
$$

The minimum follows from the contractual structure: the investor must be protected against the least favorable redemption scenario the issuer is allowed to impose.

Consider a bond with face value $1{,}000{,}000$, a nominal annual coupon rate of $10\%$ payable semiannually, and final maturity in eight years. It may be redeemed at year 4 for $1{,}040{,}000$, at year 6 for $1{,}020{,}000$, or at year 8 for $1{,}000{,}000$.

At a target semiannual yield of $4\%$, the scenario prices are approximately $1{,}096{,}555.06$, $1{,}106{,}342.68$, and $1{,}116{,}522.96$. Therefore,

$$
P_{\max}=1{,}096{,}555.06.
$$

```{r callable-bond}
call_dates <- c(4, 6)
call_values <- c(1040000, 1020000)

target_yield <- 0.04

callable_price <-
  bond_callable_price(
    face = 1000000,
    c = 0.10,
    n = 8,
    k = 2,
    call_t = call_dates,
    call_R = call_values,
    y_effective_per_period = target_yield,
    R = 1000000
  )

callable_price
```

The same contractual logic applies when price is known and yields are compared: the yield to worst is the minimum among the relevant yields.

## From price to sensitivity

For a conventional bond with positive future cash flows,

$$
P(i)
=
\sum_{t=1}^{n}CF_t(1+i)^{-t}.
$$

Differentiating,

$$
\frac{dP}{di}
=
-\sum_{t=1}^{n}tCF_t(1+i)^{-(t+1)}<0,
$$

and

$$
\frac{d^2P}{di^2}
=
\sum_{t=1}^{n}t(t+1)CF_t(1+i)^{-(t+2)}>0.
$$

The price-yield relationship is therefore decreasing and convex. These two derivatives lead naturally to duration and convexity.

## Duration and convexity of a cash-flow stream

For payments $C_j$ at times $t_j$,

$$
P(i)
=
\sum_j C_j(1+i)^{-t_j}.
$$

Define present-value weights

$$
w_j(i)
=
\frac{C_j(1+i)^{-t_j}}{P(i)}.
$$

For positive cash flows, Macaulay duration is

$$
D_{\text{Mac}}(i)
=
\sum_j t_jw_j(i),
$$

while modified duration is

$$
D_{\text{mod}}(i)
=
-\frac{1}{P(i)}\frac{dP(i)}{di}
=
\frac{D_{\text{Mac}}(i)}{1+i}.
$$

Thus,

$$
\frac{\Delta P}{P}
\approx
-D_{\text{mod}}\Delta i.
$$

Relative convexity is

$$
C(i)
=
\frac{1}{P(i)}\frac{d^2P(i)}{di^2},
$$

so the second-order approximation is

$$
\frac{\Delta P}{P}
\approx
-D_{\text{mod}}\Delta i
+
\frac12 C(\Delta i)^2.
$$

Duration gives a first-order approximation; convexity adds the effect of curvature.

## A general cash-flow example

Consider payments of 120, 160, 220, and 300 million at years 1, 2, 4, and 6, valued at an annual effective rate of $7\%$. The analytical results are approximately

$$
P_0=619.64,
\qquad
D_{\text{Mac}}=3.6512,
\qquad
D_{\text{mod}}=3.4123,
\qquad
C=18.0646.
$$

```{r cash-flow-duration-convexity}
cash_flows <- c(120, 160, 220, 300)
payment_times <- c(1, 2, 4, 6)
valuation_rate <- 0.07

macaulay_duration <-
  duration_cash_flow(
    cf = cash_flows,
    t = payment_times,
    rate = valuation_rate,
    duration = "macaulay",
    output = "value"
  )

modified_duration <-
  duration_cash_flow(
    cf = cash_flows,
    t = payment_times,
    rate = valuation_rate,
    duration = "modified",
    output = "value"
  )

cash_flow_convexity <-
  convexity_cash_flow(
    cf = cash_flows,
    t = payment_times,
    rate = valuation_rate,
    output = "value"
  )

tibble::tibble(
  measure = c("Macaulay duration", "Modified duration", "Convexity"),
  value = c(macaulay_duration, modified_duration, cash_flow_convexity)
)
```

The functions reproduce quantities already defined from the value function and its derivatives.

## Specialization to bonds

For a bond with face value $F$, annual coupon rate $c$, $k$ coupon payments per year, maturity $n$ years, and redemption value $R$,

$$
C=\frac{Fc}{k},
\qquad
N=nk.
$$

Let $j$ be the effective yield per coupon period and $y$ the annual effective yield. Then

$$
1+y=(1+j)^k,
\qquad
j=(1+y)^{1/k}-1.
$$

If $D_p$ denotes Macaulay duration in coupon periods, duration in years is

$$
D=\frac{D_p}{k}.
$$

The rate variable used in the sensitivity measure must match the rate change being studied. A duration defined with respect to $j$ must be combined with $\Delta j$; one defined with respect to $y$ must be combined with $\Delta y$.

Consider

$$
F=R=1000,
\qquad
c=8\%,
\qquad
n=10,
\qquad
k=2,
\qquad
y=9\%.
$$

```{r bond-duration-convexity}
bond_duration_result <-
  bond_duration(
    face = 1000,
    c = 0.08,
    n = 10,
    k = 2,
    y = 0.09,
    y_type = "effective"
  )

bond_convexity_result <-
  bond_convexity(
    face = 1000,
    c = 0.08,
    n = 10,
    k = 2,
    y = 0.09,
    y_type = "effective"
  )

bond_duration_result
bond_convexity_result
```

For this bond, the Macaulay duration is approximately $13.9529$ coupon periods, or $6.9764$ years.

## Portfolio duration and convexity

Portfolio sensitivities are value-weighted averages when the individual measures refer to the same rate movement.

```{r portfolio-risk}
portfolio <-
  tibble::tibble(
    asset = c("A", "B", "C"),
    units = c(450, 300, 180),
    price = c(96.20, 102.75, 88.50),
    duration = c(2.80, 5.60, 9.40),
    convexity = c(11.40, 43.20, 104.60)
  ) |>
  dplyr::mutate(
    market_value = units * price
  )

portfolio_duration_result <-
  portfolio_duration(
    portfolio,
    col_P = "market_value",
    col_D = "duration",
    col_portfolio = NULL
  )

portfolio_convexity_result <-
  portfolio_convexity(
    portfolio,
    col_P = "market_value",
    col_C = "convexity",
    col_portfolio = NULL
  )

portfolio_duration_result
portfolio_convexity_result
```

The weights are based on economic value, not simply on face amount or number of securities.

## From sensitivity to immunization

For a positive cash-flow stream and a horizon $H$, suppose all cash flows are valued and reinvested using the same flat effective rate $i$. The accumulated value at the horizon is

$$
W_H(i)
=
(1+i)^HP(i).
$$

Therefore,

$$
\frac{W_H'(i)}{W_H(i)}
=
\frac{H-D_{\text{Mac}}(i)}{1+i}.
$$

If

$$
H=D_{\text{Mac}}(i_0),
$$

then $W_H'(i_0)=0$. At a horizon equal to Macaulay duration, the local price effect and the local reinvestment effect offset each other to first order.

## The surplus function and Redington

Let

$$
S(i)=PV_A(i)-PV_L(i).
$$

If

$$
PV_A(i_0)=PV_L(i_0),
$$

then $S(i_0)=0$. If, in addition,

$$
D_A(i_0)=D_L(i_0),
$$

then $S'(i_0)=0$.

Since $PV''(i)=PV(i)C(i)$, under present-value equality,

$$
S''(i_0)
=
PV_A(i_0)\left[C_A(i_0)-C_L(i_0)\right].
$$

Thus the classical Redington conditions are

$$
PV_A(i_0)=PV_L(i_0),
$$

$$
D_A(i_0)=D_L(i_0),
$$

and

$$
C_A(i_0)>C_L(i_0).
$$

Redington is therefore a local curvature condition. By contrast, duration-convexity matching imposes $C_A=C_L$, so $S''(i_0)=0$. These are different constructions.

## An integrated immunization example

Consider liabilities

$$
L=(30{,}000,\ 40{,}000,\ 50{,}000)
$$

payable at years $3$, $5$, and $8$, valued at an annual effective rate of $5\%$. Their present value is approximately $91{,}098.14$, with Macaulay duration $5.5455$ and convexity $36.7184$.

Suppose zero-coupon assets with redemption value 1,000 are available at maturities 2, 5, and 10 years.

```{r immunization-inputs}
liabilities <- c(30000, 40000, 50000)
liability_times <- c(3, 5, 8)
reference_rate <- 0.05

asset_maturities <- c(2, 5, 10)
asset_redemption <- 1000

asset_prices <-
  asset_redemption *
  (1 + reference_rate)^(-asset_maturities)

asset_durations <-
  asset_maturities

asset_convexities <-
  asset_maturities *
  (asset_maturities + 1) /
  (1 + reference_rate)^2
```

Using the 2-year and 10-year assets:

```{r duration-match}
duration_match <-
  immunize_duration(
    L = liabilities,
    t = liability_times,
    P = asset_prices[c(1, 3)],
    D = asset_durations[c(1, 3)],
    i = reference_rate,
    i_type = "effective"
  )

duration_match
```

`immunize_duration()` matches present value and duration. It does not by itself prove Redington; the convexity inequality must be checked separately.

With all three assets, present value, duration, and convexity can be matched simultaneously:

```{r duration-convexity-match}
duration_convexity_match <-
  immunize_duration_convexity(
    L = liabilities,
    t = liability_times,
    P = asset_prices,
    D = asset_durations,
    C = asset_convexities,
    i = reference_rate,
    i_type = "effective"
  )

duration_convexity_match
```

`immunize_duration_convexity()` should be interpreted exactly this way: it matches present value, duration, and convexity. It is not a direct implementation of Redington and it does not establish full immunization.

## The immunization gap

The surplus function has a direct interpretation:

$$
S(i)>0 \Rightarrow \text{asset surplus},
$$

$$
S(i)=0 \Rightarrow \text{exact value equality},
$$

$$
S(i)<0 \Rightarrow \text{asset insufficiency}.
$$

`plot_immunization_gap()` can be used to inspect this gap over a finite range of rates.

```{r immunization-gap, fig.cap="Immunization gap around the reference rate."}
asset_cash_flows <- list(
  list(
    cf = 444998.22,
    t = 3
  ),
  list(
    cf = 561800.00,
    t = 7
  )
)

plot_immunization_gap(
  L = 1000000,
  t = 5,
  asset_cashflows = asset_cash_flows,
  w = c(1, 1),
  i = 0.06,
  i_type = "effective",
  delta = 0.02
)
```

The graph is a diagnostic tool. A finite grid of rates can illustrate the shape of $S(i)$, but it cannot prove $S(i)\ge0$ for every admissible rate. Full immunization requires a stronger structural argument.

## Final actuarial reading

The valuation of a fixed-income position begins with cash flows and a rate basis. Duration and convexity then describe how that value responds locally to changes in the rate.

The sequence is

$$
P(i)
\longrightarrow
P'(i)
\longrightarrow
D
\longrightarrow
P''(i)
\longrightarrow
C,
$$

followed by

$$
PV_A-PV_L
\longrightarrow
\text{duration matching}
\longrightarrow
\text{immunization}.
$$

Throughout the vignette, `tidyactuarial` enters after the actuarial structure has been identified. It reproduces bond values, yields, sensitivities, portfolio measures, and immunization systems while the mathematical interpretation remains explicit.

The same precision applies to the immunization functions: `immunize_duration()` matches present value and duration; `immunize_duration_convexity()` matches present value, duration, and convexity; Redington additionally requires favorable curvature; and full immunization requires a stronger structural argument.

The software verifies the equations being solved. It does not replace the conditions that define the actuarial strategy.
